Reagents: expressing and composing fine-grained concurrency

### 2011-11-07 paper/paper.tex
submittable version

@@

\begin{abstract}
blah blah\\
blah blah\\
blah blah\\
blah blah\\
blah blah\\
blah blah
\end{abstract}

@@

\hspace{\stretch{1}}\emph{Programs are what happens between cache
  misses}.\footnote{Folklore; origin unknown.}

@@

When \lstinline{push} or \lstinline{pop} retry, they follow an
exponential backoff pattern with \emph{spinwaiting} for elimination
offers, exactly as described in \secref{background}.  The backoff
logic follows directly from the implementation of reagents; see
\secref{implementation} for details.

@@

\elide{
def !(a: A): B = tryReact(a, empty, null) match {
  case (_: Failure) => {
    val backoff = new Backoff
    def retryLoop(shouldBlock: Boolean): B = {
      val offer = if (maySync || shouldBlock) 
                    new Offer[B]
                  else null
      tryReact(a, empty, offer) match {
        case (f: Failure) => {
          if (f == Block && offer != null) 
            LockSupport.park 
          else backoff.once

          // rescind offer, but check for completion
          if (offer != null && offer.tryAbort)
            offer.answer
          else
            retryLoop(f == Block)
        }
        case ans   => ans.asInstanceOf[B]
      }
    }
    backoff.once
    retryLoop(false)
  }
  case ans => ans.asInstanceOf[B]
}
}

@@

In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
whose subclasses are hidden.  The subclasses roughly correspond to the
combinator functions (which are responsible for instantiating them), and
instances of the subclasses store the arguments given to their corresponding
combinator.  Each subclass provides an implementation of the
\lstinline{tryReact} method, which is an abstract method of
\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
  programming aficionados should regard \lstinline{tryReact} as a kind of
  Kleisli arrow for reagents.}
\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
\end{lstlisting}
The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
represent a union of the type \lstinline{B} with the type
\lstinline{Failure}, where the latter has just two singleton instances,
\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
transient failures.

@@

\begin{figure}[t]
\begin{lstlisting}[frame=single]
class Commit[A] extends Reagent[A,A] {
  def tryReact(a: A, rx: Reaction, offer: Offer[A]) = 
    if (offer != null && !offer.rescind) offer.answer
    else if (rx.commit) a 
    else Retry
}
\end{lstlisting}
\vskip -21pt
\begin{lstlisting}[frame=single]
class CAS[A,R](ref: Ref[A], ov: A, nv: A, k: Reagent[A,R]) 
  extends Reagent[Unit,R] {
  def tryReact(u: Unit, rx: Reaction, offer: Offer[R]) = 
    if (!rx.hasCAS && !k.hasCAS)
      if (ref.cas(ov, nv)) k.tryReact((), rx, offer)
      else Retry
    else
      k.tryReact((), rx.withCAS(ref, ov, nv), offer)
}
\end{lstlisting}
\vskip -21pt
\begin{lstlisting}[frame=single]
class Choice[A,B](r1: Reagent[A,B], r2: Reagent[A,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    r1.tryReact(a, rx, offer) match {
    case Retry => r2.tryReact(a, rx, offer) match {
      case (_: Failutre) => Retry // must retry r1
      case ans           => ans
    }
    case Block => r2.tryReact(a, rx, offer)
    case ans   => ans
  }
}
\end{lstlisting}
\vskip -21pt
\begin{lstlisting}[frame=single]
class Computed[A,B](c: PartialFunction[A, Reagent[Unit,B]]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    if (c.isDefinedAt(a)) c(a).tryReact((), rx, offer)
    else Block
}
\end{lstlisting}
\vskip -21pt
\begin{lstlisting}[frame=single]
class Swap[A,B,R](ep: Endpoint[A,B], k: Reagent[B, R]) 
extends Reagent[A,R] {
  def tryReact(a: A, rx: Reaction, offer: Offer[R]) = {
    if (offer != null) // send message if so requested
      ep.put(new Message(a, rx, k, offer))
    def tryFrom(cur: Cursor, failMode: Failure): Any = 
      cur.getNext match {
        case Some(msg, next) =>
          msg.exchange(k).tryReact(a, rx, offer) match {
            case Retry => tryFrom(next, Retry)
            case Block => tryFrom(next, failMode)
            case ans   => ans
          }
        case None => failMode
      }    
    tryFrom(ep.dual.cursor, Retry) // attempt reaction
  }
\end{lstlisting}
\nocaptionrule
\caption{Excerpts from core reagent subclasses}
\label{fig:impl}
\end{figure}

@@

\begin{figure*}
\newcommand{\gscale}{0.72}
\hspace{-30pt}
\begin{tabular}[t]{l@{\hspace{2mm}}c}
\begin{sideways}
\hspace{0.5in}
\textbf{Throughput}: \quad
  iterations/$\mu$s \quad
  (bigger is better)
\end{sideways} &
\begin{tabular}[b]{l}
\includegraphics[scale=\gscale]{graphs/PushPop-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/PushPop-1000.pdf}
\hspace{4mm}
\includegraphics[scale=\gscale]{graphs/StackTransfer-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/StackTransfer-1000.pdf}
\\
\\
\includegraphics[scale=\gscale]{graphs/EnqDeq-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/EnqDeq-1000.pdf}
\hspace{4mm}
\includegraphics[scale=\gscale]{graphs/QueueTransfer-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/QueueTransfer-1000.pdf}
\end{tabular}
\\
& \textbf{Threads}\quad (on 16-way machine) \\
& 
\includegraphics[scale=0.9]{graphs/reagent-based.pdf}
\qquad\qquad
\includegraphics[scale=0.9]{graphs/hand-built.pdf}
\qquad\qquad
\includegraphics[scale=0.9]{graphs/lock-based.pdf}
\qquad\qquad
\includegraphics[scale=0.9]{graphs/stm-based.pdf}
\\
\end{tabular}
\nocaptionrule
\caption{Benchmarking results}
\label{fig:perf}
\end{figure*}

@@

Haskell's STM~\cite{Harris2005a} was inspirational in showing that
transactions can be represented via monads~\cite{PeytonJones1993}, explicitly
composed, and combined with blocking and choice operators.  Reagents also form
a monad, but we have chosen an interface closer to
\emph{arrows}~\cite{Hughes2000}, to encourage static reagent layout wherever
possible~(\secref{dynamic}).  Like \lstinline{orElse} in Haskell's STM, our
choice operator is left-biased.  But unlike \lstinline{orElse}, our choice
operator will attempt the right-hand side even when the left-hand side has
only failed transiently (rather than permanently).  While the distinction may
appear somewhat technical, it is crucial for supporting examples like the
elimination backoff stack~(\secref{choice}).

@@

A central tenet of transactions is \emph{isolation}: transactions should not
be aware of concurrent changes to memory from other transactions.  But
sometimes it is desirable for concurrent transactions to coordinate or
otherwise communicate while executing.  Two very recent papers have proposed
mechanisms for incorporating communication with STM, in the form of
Communicating Transactions~\cite{Lesani2011} and Transaction
Communicators~\cite{Luchangco2011b} (the latter extends earlier work on
Transaction Synchronizers~\cite{Luchangco2005}).  In both cases, a key
question is how the expected isolation of shared memory can safely coexist
with concurrent communication.  Communicating Transactions use explicit,
asynchronous message passing to communicate; the mechanism is entirely
separate from shared memory, which retains isolation.  On the other hand,
Transaction Communicators allow isolation to be weakened in a controlled way.

Our mixture of message-passing and shared-state combinators most closely
resembles the work on Communicating Transactions.  Of course, the most
important difference is in the way we deal with shared state (and the
guarantees provided thereby), which we discussed in~\secref{stm}.  Aside from
that, we believe that \emph{synchronous} communication is a better choice for
expressing patterns like elimination~(\secref{choice}), since the pattern
relies on both participants being mutually-aware.  %TODO: say more here

>>>>

%\usepackage[compact]{titlesec}

@@

\begin{abstract}
Efficient communication and synchronization is crucial for fine-grained
parallelism.  Libraries providing such features, while indispensable, are
difficult to write, and often cannot be tailored or composed to meet the needs
of specific users.  We introduce \emph{reagents}, a set of combinators for
concisely expressing concurrency algorithms.  Reagents scale as well as their
hand-coded counterparts, while providing the composability existing libraries
lack.
\end{abstract}

@@

\hspace{\stretch{1}}\emph{Programs are what happens between cache
  misses}. %{\small[origin unknown]}%\footnote{Folklore; origin unknown.}

@@

When \lstinline{push} or \lstinline{pop} retry, they follow an
exponential backoff pattern with spinwaiting for elimination
offers\elide{, exactly as described in \secref{background}}.  The backoff
logic follows directly from the implementation of reagents~(\secref{implementation}).

@@

@@

In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
whose subclasses are hidden.  The subclasses roughly correspond to the
combinator functions (which are responsible for instantiating them), and
instances of the subclasses store the arguments given to their corresponding
combinator.  Each subclass provides an implementation of the
\lstinline{tryReact} method, which is an abstract method of
\lstinline{Reagent[A,B]} with the following signature:\elide{\footnote{ Functional
  programming aficionados should regard \lstinline{tryReact} as a kind of
  Kleisli arrow for reagents.}}
\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
\end{lstlisting}
The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
represent a union of the type \lstinline{B} with the type
\lstinline{Failure}, where the latter has just two singleton instances,
\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
transient failures.

@@

\begin{figure}[t]
\begin{lstlisting}[frame=single]
class Commit[A] extends Reagent[A,A] {
  def tryReact(a: A, rx: Reaction, offer: Offer[A]) = 
    if (offer != null && !offer.rescind) offer.answer
    else if (rx.commit) a 
    else Retry
}
\end{lstlisting}
\vskip -4pt
\begin{lstlisting}[frame=single]
class CAS[A,R](ref: Ref[A], ov: A, nv: A, k: Reagent[A,R]) 
  extends Reagent[Unit,R] {
  def tryReact(u: Unit, rx: Reaction, offer: Offer[R]) = 
    if (!rx.hasCAS && !k.hasCAS)
      if (ref.cas(ov, nv)) k.tryReact((), rx, offer)
      else Retry
    else
      k.tryReact((), rx.withCAS(ref, ov, nv), offer)
}
\end{lstlisting}
\vskip -4pt
\begin{lstlisting}[frame=single]
class Choice[A,B](r1: Reagent[A,B], r2: Reagent[A,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    r1.tryReact(a, rx, offer) match {
      case Retry => r2.tryReact(a, rx, offer) match {
        case (_: Failutre) => Retry // must retry r1
        case ans           => ans
      }
      case Block => r2.tryReact(a, rx, offer)
      case ans   => ans
    }
}
\end{lstlisting}
\vskip -4pt
\begin{lstlisting}[frame=single]
class Computed[A,B](c: PartialFunction[A, Reagent[Unit,B]]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    if (c.isDefinedAt(a)) c(a).tryReact((), rx, offer)
    else Block
}
\end{lstlisting}
\vskip -4pt
\begin{lstlisting}[frame=single]
class Swap[A,B,R](ep: Endpoint[A,B], k: Reagent[B, R]) 
extends Reagent[A,R] {
  def tryReact(a: A, rx: Reaction, offer: Offer[R]) = {
    if (offer != null) // send message if so requested
      ep.put(new Message(a, rx, k, offer))
    def tryFrom(cur: Cursor, failMode: Failure): Any = 
      cur.getNext match {
        case Some(msg, next) =>
          msg.exchange(k).tryReact(a, rx, offer) match {
            case Retry => tryFrom(next, Retry)
            case Block => tryFrom(next, failMode)
            case ans   => ans
          }
        case None => failMode
      }    
    tryFrom(ep.dual.cursor, Retry) // attempt reaction
  }
\end{lstlisting}
\nocaptionrule
\caption{Excerpts from core reagent subclasses}
\label{fig:impl}
\end{figure}

@@

\begin{figure*}
\newcommand{\gscale}{0.72}
\hspace{-30pt}
\begin{tabular}[t]{l@{\hspace{2mm}}c}
\begin{sideways}
\hspace{0.5in}
\textbf{Throughput}: \quad
  iterations/$\mu$s \quad
  (bigger is better)
\end{sideways} &
\begin{tabular}[b]{l}
\includegraphics[scale=\gscale]{graphs/PushPop-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/PushPop-1000.pdf}
\hspace{4mm}
\includegraphics[scale=\gscale]{graphs/StackTransfer-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/StackTransfer-1000.pdf}
\\
\\
\includegraphics[scale=\gscale]{graphs/EnqDeq-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/EnqDeq-1000.pdf}
\hspace{4mm}
\includegraphics[scale=\gscale]{graphs/QueueTransfer-100.pdf}
\hspace{1mm}
\includegraphics[scale=\gscale]{graphs/QueueTransfer-1000.pdf}
\end{tabular}
\\
& \textbf{Threads}\quad (on 16-way machine) \\
\\
& 
\includegraphics[scale=0.9]{graphs/reagent-based.pdf}
\qquad\qquad
\includegraphics[scale=0.9]{graphs/hand-built.pdf}
\qquad\qquad
\includegraphics[scale=0.9]{graphs/lock-based.pdf}
\qquad\qquad
\includegraphics[scale=0.9]{graphs/stm-based.pdf}
\\
\end{tabular}
\nocaptionrule
\caption{Benchmarking results}
\label{fig:perf}
\end{figure*}

@@

Haskell's STM~\cite{Harris2005a} was inspirational in showing that transactions
can be represented via monads~\cite{PeytonJones1993}, explicitly composed, and
combined with blocking and choice operators.  Reagents also form a monad, but we
have chosen an interface closer to \emph{arrows}~\cite{Hughes2000}, to encourage
static reagent layout wherever possible~(\secref{dynamic}).  Like
\lstinline{orElse} in Haskell's STM, our choice operator is left-biased.  But
unlike \lstinline{orElse}, our choice operator will attempt the right-hand side
even when the left-hand side has only failed transiently (rather than
permanently).  While the distinction appears technical, it is crucial for
supporting examples like the elimination backoff stack~(\secref{choice}).

@@

A central tenet of transactions is \emph{isolation}: transactions should not
be aware of concurrent changes to memory from other transactions.  But
sometimes it is desirable for concurrent transactions to coordinate or
otherwise communicate while executing.  Two very recent papers have proposed
mechanisms for incorporating communication with STM, in the form of
Communicating Transactions~\cite{Lesani2011} and Transaction
Communicators~\cite{Luchangco2011b}\elide{ (the latter extends earlier work on
Transaction Synchronizers~\cite{Luchangco2005})}.  In both cases, a key
question is how the expected isolation of shared memory can safely coexist
with concurrent communication.  Communicating Transactions use explicit,
asynchronous message passing to communicate; the mechanism is entirely
separate from shared memory, which retains isolation.  On the other hand,
Transaction Communicators allow isolation to be weakened in a controlled way.

Our mixture of message-passing and shared-state combinators most closely
resembles the work on Communicating Transactions.  Of course, the most important
difference is in the way we deal with shared state \elide{(and the guarantees
  provided thereby), which we} discussed in~\secref{stm}.  We also
believe that \emph{synchronous} communication is better for expressing
patterns like elimination~(\secref{choice}), since they rely on 
participants being mutually-aware.  %TODO: say more here
